---
title: "Factors Influencing Housing Values in Boston"
subtitle: "A Exploratory and Regression Analysis"
author: |
  Xiaoying Ma  
  
date: "06/09/2026"
output:
  html_document:
    theme: yeti
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    highlight: tango
    #css: styles.css
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: true
    fig_width: 8
    fig_height: 6
    latex_engine: xelatex
  word_document:
    toc: true
    toc_depth: 4
    number_sections: true
abstract: |
  This report analyzes the factors that influence housing values in the Boston area. The main response variable is **MEDV**, the median value of owner-occupied homes in thousands of dollars. The analysis uses data inspection, distributional exploration, location-related visualization, feature engineering, and regression modeling to study how neighborhood, environmental, accessibility, and river-adjacent characteristics are associated with housing value.

  
---


```{css echo = FALSE}
/* ============ CSS Styling for Lecture Notes ============ */

body {
  font-family: 'Segoe UI', 'Inter', system-ui, -apple-system, sans-serif;
  line-height: 1.7;
  font-size: 1.4em;
  color: #2d1b29;
  
  background-color: #fdfafd;
}

#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.4em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
    /*font-weight: bold; */
    font-size: 1.3rem;
    font-weight: 500;
    color: #6E3061;
}

.title {
  font-size: 2.3rem;
  color: #6E3061;
  text-align: left;
}

.subtitle {
  font-size: 2.1rem;
  color: #6E3061;
  text-align: left;
  font-weight: 800;
}

.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 16px;
  /*font-family: "Times New Roman", Times, serif;*/
  color: #6E3061;
  text-align: left;
  font-weight: 400;
}

.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 14px;
  /*font-family: "Times New Roman", Times, serif;*/
  color: #6E3061;
  text-align: left;
  font-weight: 400;
}


h1 {
  color: #6E3061;
  border-bottom: 3px solid #6E3061;
  padding-bottom: 8px;
  margin-top: 40px;
  font-weight: bold;
  font-size: 2.2rem;
}

h2 {
  color: #6E3061;
  border-left: 5px solid #6E3061;
  padding-left: 12px;
  margin-top: 30px;
  font-weight: bold;
  font-size: 1.8rem;
}

h3 {
  color: #8b4a7d;
  margin-top: 24px;
  font-weight: bold;
  font-size: 1.8m;
}

h4 {
  color: #a05f92;
  margin-top: 18px;
  font-weight: bold;
  font-size: 1.4rem;
}

code {
  background-color: #f4edf3;
  color: #4a2040;
  padding: 2px 6px;
  border-radius: 4px;
  font-family: 'Fira Code', 'JetBrains Mono', 'Consolas', monospace;
  font-size: 0.9em;
  border: 1px solid #d4c0d1;
}

pre {
  background-color: #f9f5f9;
  border: 1px solid #d4c0d1;
  border-left: 4px solid #6E3061;
  border-radius: 6px;
  padding: 14px;
}

pre code {
  background-color: transparent;
  border: none;
  padding: 0;
}

blockquote {
  background-color: #f4edf3;
  border-left: 4px solid #6E3061;
  padding: 12px 18px;
  margin: 16px 0;
  border-radius: 0 6px 6px 0;
  color: #4a2040;
}

.note-box {
  background-color: #f0e6ee;
  border: 1px solid #c49bb8;
  border-radius: 8px;
  padding: 16px 20px;
  margin: 18px 0;
}

.note-box strong {
  color: #6E3061;
}

.tip-box {
  background-color: #e8f5e9;
  border: 1px solid #81c784;
  border-radius: 8px;
  padding: 16px 20px;
  margin: 18px 0;
}

.tip-box strong {
  color: #2e7d32;
}

.warning-box {
  background-color: #fff3e0;
  border: 1px solid #ffb74d;
  border-radius: 8px;
  padding: 16px 20px;
  margin: 18px 0;
}

.warning-box strong {
  color: #e65100;
}

table {
  width: 100%;
  border-collapse: collapse;
  margin: 16px 0;
}

table th {
  background-color: #6E3061;
  color: white;
  padding: 10px 12px;
  text-align: left;
  font-weight: 600;
}

table td {
  padding: 8px 12px;
  border-bottom: 1px solid #d4c0d1;
}

table tr:nth-child(even) td {
  background-color: #faf5f9;
}

table tr:hover td {
  background-color: #f0e6ee;
}

.toc {
  background-color: #faf5f9;
  border: 1px solid #d4c0d1;
  border-radius: 8px;
  padding: 16px 20px;
  margin-bottom: 24px;
}

.figure {
  border: 1px solid #e0d1db;
  border-radius: 8px;
  padding: 8px;
  margin: 16px 0;
  box-shadow: 0 2px 8px rgba(110, 48, 97, 0.08);
}
```


```{r setup, include=FALSE}
# =============================
#  SETUP AND CONFIGURATION
# =============================

# ---- Knitr Global Options ----
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 7,
  dpi = 150,
  cache = FALSE,
  comment = NA,
  out.width = "100%",
  fig.pos = "H"
)

# ---- R Global Options ----
options(
  scipen = 999,
  digits = 4,
  width = 100,
  dplyr.summarise.inform = FALSE,
  stringsAsFactors = FALSE,
  max.print = 100
)
```



```{r echo = FALSE, message=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Quick install and load for R Markdown
packages <- c("tidyverse", "plotly", "leaflet", "DT", "corrplot",
              "GGally", "MASS", "patchwork", "scales",
              "viridis", "RColorBrewer", "ggridges", "gghighlight",
              "ggrepel", "cowplot", "sf", "ggspatial", "knitr",
              "kableExtra", "gridExtra", "leaps", "ggfortify", "pander")

# Install missing packages
missing_pkgs <- packages[!(packages %in% rownames(installed.packages()))]

if(length(missing_pkgs) > 0){
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))


# ---- Global Theme Configuration ----
my.theme = function(){
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = "#2c3e50", hjust = 0.5),
      plot.subtitle = element_text(color = "#7f8c8d", size = 11),
      plot.caption = element_text(color = "#95a5a6", size = 8, face = "italic"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#ecf0f1", linewidth = 0.3),
      axis.title = element_text(size = 11, color = "#34495e"),
      axis.text = element_text(size = 10, color = "#2c3e50"),
      strip.text = element_text(face = "bold", size = 11),
      strip.background = element_rect(fill = "#f8f9fa", color = NA)
    )
 }
```

# Introduction: 

The **Boston Housing Dataset** contains housing and neighborhood information for census tracts in the Boston Standard Metropolitan Statistical Area. The data include median housing value, crime rate, environmental measures, accessibility measures, school-related variables, and neighborhood characteristics. The response variable selected for regression analysis is **MEDV**, the median value of owner-occupied homes in thousands of dollars.
 


**Objectives**

This project aims to achieve the following objectives:

* Identify the key factors that significantly influence housing prices in Boston

* Quantify the magnitude and direction of each predictor’s effect on median home value

* Visualize spatial patterns and relationships among housing, environmental, and socioeconomic variables

* Develop a predictive model for estimating Boston housing prices

* Demonstrate the application of the Visualization Framework in housing market analysis


# Data Acquisition and Processing

This dataset contains housing and neighborhood information for census tracts in the Boston metropolitan area. It includes structural, environmental, accessibility, and socioeconomic variables, while the response variable `medv` records the median value of owner-occupied homes in thousands of dollars.


```
╔══════════════════════════════════════════════════════════════╗
║              DATA ACQUISITION SUMMARY                        ║
╠══════════════════════════════════════════════════════════════╣
║  Dataset:    Boston Housing Dataset                          ║
║  Source:     Kaggle                                          ║
║  Period:     1978                                            ║
║  Geography:  Boston Metropolitan Area                        ║
╚══════════════════════════════════════════════════════════════╝
```


* **Variable Dictionary in the Boston Housing Dataset **

| Variable Name |   Description |
|:------------------|:----------------------------------------------------------|
|  CRIM |  Per capita crime rate by town . | 
|  ZN|  Proportion of residential land zoned for lots over 25,000 sq.ft.| 
|  INDUS|  Proportion of non-retail business acres per town.|  
|  CHAS|   Charles River dummy variable (1 if tract bounds river; 0 otherwise). | 
|  NOX|  Nitric oxides concentration (parts per 10 million). | 
|  RM|  Average number of rooms per dwelling. | 
|  AGE|  Proportion of owner-occupied units built prior to 1940. | 
|  DIS|  Weighted distances to five Boston employment centres . | 
|  RAD|   Index of accessibility to radial highways. | 
|  TAX|   Full-value property-tax rate per $10,000 . |
|  PTRATIO|   Pupil-teacher ratio by town . | 
|  B|  1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town. | 
|  LSTAT|  % lower status of the population. | 
|  MEDV|  Median value of owner-occupied homes in $1000's. | 
 

* **Initial Data Processing**

The initial data processing includes checking for missing values and preparing variable types for later analysis. Since the Boston housing data mainly contain numeric variables, only basic cleaning and simple variable transformation are needed.

```{r}
# ---- Download / Load Dataset ----
library(MASS)
housing_raw <- Boston

# Check for missing values and basic structure
missing_summary <- summary(housing_raw)
print(missing_summary)

```

There are no missing values in this data set. For illustrative purposes, we still show a simple cleaning step by removing observations with missing components if any were present. 

```{r import-data}
# Read the Boston Housing analytic dataset.
# Put this Rmd file and BostonHousing_analytic_dataset.csv in the same folder before knitting.
boston <- read.csv("BostonHousing_analytic_dataset.csv")

# Inspect the first few observations.
head(boston)
```

```{r data-structure}
# ======================================= 
#   DATA CLEANING AND TRANSFORMATION
# ======================================= 

# ---- Initial Cleaning ----
housing_clean <- housing_raw %>%
  # Remove observations with missing target variable
  filter(!is.na(medv)) %>%      # no missing in this data set. Use this
                                # example to demonstrate the use of filter()
  
  # Convert selected variables for later modeling/visualization
  mutate(
    chas_factor = as.factor(chas)   # Charles River dummy variable as factor
  )
```

# Visual Exploratory Data Analysis


The above descriptive summary indicates that there are no missing values in the data set. In the subsequent sections, we inspect variables and provide recommendations on feature engineering.



## Distributional Patterns of Features

We examine the distributions of categorical and numerical variables to identify predictors that may require regrouping or transformation.

* **Categorical Variables**

There is one main categorical variable in the Boston housing dataset, namely `chas`, which indicates whether the tract bounds the Charles River (`1 = yes`, `0 = no`). Since this is a binary variable, there is no issue of sparse categories.

```{r}
# Convert CHAS to factor for plotting and interpretation
housing_clean$chas_factor <- factor(housing_clean$chas,
                                    levels = c(0, 1),
                                    labels = c("No", "Yes"))

ggplot(housing_clean, aes(x = chas_factor)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5, size = 4) +
  labs(title = "Distribution of Charles River Indicator",
       x = "Bounds Charles River",
       y = "Frequency") +
  my.theme()
```

The above bar chart does not indicate a serious imbalance problem for this binary variable. The relationship between `chas` and median house value will be examined later.


* **Numerical Variables**

The Boston housing dataset contains several numerical predictors. In this project, we focus on four important continuous variables: average number of rooms (`rm`), percentage of lower status population (`lstat`), nitrogen oxides concentration (`nox`), and weighted distance to employment centers (`dis`). The following pairwise plot shows their marginal distributions and mutual relationships.

```{r}

# Define panel functions
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use = "complete.obs")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste(prefix, txt, sep = "")
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * 0.8, col = "steelblue", font = 1.1)
}

panel.hist <- function(x, ...) {
  usr <- par("usr")
  on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5))
  h <- hist(x, plot = FALSE, breaks = 20)
  breaks <- h$breaks
  nB <- length(breaks)
  y <- h$counts
  y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col = "steelblue", border = "darkblue")
}

panel.scatter <- function(x, y, ...) {
  points(x, y, pch = 19, col = rgb(0.3, 0.5, 0.8, alpha = 0.5), cex = 0.8)
  abline(lm(y ~ x), col = "darkred", lwd = 2)
}

# Create the pairs plot
pairs(housing_clean[, c("rm", "lstat", "nox", "dis")],
      lower.panel = panel.scatter,
      upper.panel = panel.cor,
      diag.panel = panel.hist,
      main = "Pair-wise Analysis of Selected Numerical Variables",
      cex.labels = 1.2,
      font.labels = 2)
```

The continuous variables show different distributional patterns, and several relationships are evident in the pairwise plot. For example, `rm` tends to be negatively associated with `lstat`, while `nox` and `dis` show a negative relationship. These patterns are useful for understanding the structure of the data and for selecting variables in the regression analysis later.

* **Extract Geo-spatial Information**

The Boston housing dataset does not contain exact longitude and latitude coordinates, so a geographic map cannot be constructed directly. Instead, the dataset provides several location-related variables that capture neighborhood structure, such as `dis` (weighted distance to employment centers), `rad` (accessibility to radial highways), and `chas` (whether the tract bounds the Charles River). These variables provide indirect spatial information.

```{r}
# Load required libraries
library(plotly)
library(dplyr)

# Create interactive scatter plot for Boston housing data
plot_ly(housing_clean,
        x = ~rad,
        y = ~dis,
        type = "scatter",
        mode = "markers",
        marker = list(
          color = "purple",
          size = 7,
          opacity = 0.6
        ),
        text = ~paste(
          "<b>🏠 Boston Housing Details</b><br>",
          "━━━━━━━━━━━━━━━━━━━<br>",
          "🌊 Charles River:", chas, "<br>",
          "🏘️ Average Rooms:", round(rm, 2), "<br>",
          "🌫️ NOX Concentration:", round(nox, 3), "<br>",
          "🚗 Highway Accessibility (rad):", rad, "<br>",
          "🏢 Distance to Employment Centers:", round(dis, 2), "<br>",
          "👨‍👩‍👧 Lower Status Population:", round(lstat, 2), "%<br>",
          "💰 Median House Value:", medv
        ),
        hoverinfo = "text") %>%
  layout(
    title = "Boston Housing: Accessibility vs Distance",
    xaxis = list(title = "Accessibility to Radial Highways (rad)"),
    yaxis = list(title = "Distance to Employment Centers (dis)")
  )
```

The above plot suggests that location-related variables vary systematically across neighborhoods. In particular, tracts with different levels of highway accessibility tend to have different distances to employment centers. This indicates that location may play an important role in explaining variation in housing prices.

```{r}
ggplot(housing_clean, aes(x = chas_factor, y = medv, fill = chas_factor)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Median Housing Value by Charles River Indicator",
       x = "Bounds Charles River",
       y = "Median Housing Value (medv)") +
  guides(fill = "none") +
  my.theme()
```

The boxplot shows that tracts bordering the Charles River may have a different distribution of housing values than those that do not. This suggests that location-related information in the Boston dataset captures meaningful neighborhood heterogeneity and should be considered in the subsequent modeling process.

# Feature Engineering

Based on the exploratory analysis above, several variables in the Boston housing dataset may benefit from transformation or regrouping to improve interpretability in subsequent modeling. In particular, some predictors show skewed distributions, while others are naturally suited for categorical treatment.

We perform feature engineering tasks suggested by the previous exploratory data analysis. The relationship between the engineered features and median housing value will be used for descriptive interpretation in the later analysis.

```{r}
library(MASS)
library(dplyr)

# Load Boston housing data
housing_raw <- Boston

# Initial cleaning
housing_clean <- housing_raw %>%
  filter(!is.na(medv)) %>%
  mutate(
    chas_factor = factor(chas, levels = c(0, 1), labels = c("No", "Yes"))
  )

# ======================================= 
#   FEATURE ENGINEERING
# ======================================= 

# Discretizing nitric oxides concentration
housing_clean$nox_cat <- cut(
  housing_clean$nox,
  breaks = quantile(housing_clean$nox, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Low", "Medium", "High")
)

# Discretizing weighted distance to employment centers
housing_clean$dis_cat <- cut(
  housing_clean$dis,
  breaks = quantile(housing_clean$dis, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Short", "Medium", "Long")
)

# Location-related grouping based on highway accessibility
housing_clean$location <- ifelse(
  housing_clean$rad >= median(housing_clean$rad, na.rm = TRUE),
  "urban_access",
  "other_area"
)

# Convert to factor
housing_clean$location <- as.factor(housing_clean$location)

# Analytic data
var.name <- c("medv", "chas_factor", "nox_cat", "dis_cat", "location")
FinalAnalyticData <- housing_clean[, var.name]
```

## Relationship Between Features and the House Value

After conducting visual exploratory data analysis, we now examine whether the engineered predictor variables are associated with median house value. Since this part is intended as descriptive exploration only, grouped boxplots are used to assess these relationships.

```{r}
# Read the analytic data
df <- FinalAnalyticData

# Convert categorical variables to factors with proper levels
df$chas_factor <- factor(df$chas_factor,
                         levels = c("No", "Yes"))

df$nox_cat <- factor(df$nox_cat,
                     levels = c("Low", "Medium", "High"),
                     ordered = TRUE)

df$dis_cat <- factor(df$dis_cat,
                     levels = c("Short", "Medium", "Long"),
                     ordered = TRUE)

df$location <- factor(df$location,
                      levels = c("other_area", "urban_access"))

# 1. Relationship with chas_factor
p1 <- ggplot(df, aes(x = chas_factor, y = medv, fill = chas_factor)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "House Value by Charles River Indicator",
       x = "Bounds Charles River",
       y = "Median House Value (medv)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#2E86AB", "#A23B72")) +
  my.theme()

p1

# 2. Relationship with nox_cat
p2 <- ggplot(df, aes(x = nox_cat, y = medv, fill = nox_cat)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "House Value by NOX Category",
       x = "NOX Category",
       y = "Median House Value (medv)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_viridis_d() +
  my.theme()

p2

# 3. Relationship with dis_cat
p3 <- ggplot(df, aes(x = dis_cat, y = medv, fill = dis_cat)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "House Value by Distance to Employment Centers",
       x = "Distance Category",
       y = "Median House Value (medv)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "YlOrRd") +
  my.theme()

p3

# 4. Relationship with location
p4 <- ggplot(df, aes(x = location, y = medv, fill = location)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "House Value by Location Type",
       x = "Location Type",
       y = "Median House Value (medv)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#FF6B6B", "#4ECDC4")) +
  my.theme()

p4

# Arrange plots on a grid
if(require(gridExtra)) {
  grid.arrange(p1, p4, ncol = 2, nrow = 1)
}
```

As shown in the above plots, the engineered predictor variables appear to be associated with median house value. In particular, housing values vary across pollution categories, distance-to-employment categories, and the location-related grouping. The Charles River indicator also shows some difference in the distribution of house values. These patterns suggest that the engineered features may be useful in the subsequent regression modeling.

# Factors Influencing House Prices - Regression Analysis

This section focuses on the regression analysis of Boston housing values. The main objective is to identify predictor variables that are both practically meaningful and statistically significant in explaining variation in median house value.

We follow the general regression workflow by first identifying practically important variables to be included in the model, regardless of their statistical significance, and then defining a more comprehensive model for comparison. Since the analytic dataset contains four engineered predictors, we define the full model to include all four. The final candidate model will be selected based on regression results and model diagnostics.

To identify the final model, we will examine standard regression diagnostics and apply suitable remedies when necessary.

## Candidate Models

The following two base models are used to search the potential optimal candidate model:

1. **Small model** contains only `location`:

$$
\text{MEDV} = \alpha_0 + \alpha_1 \times \text{location}
$$
```{r}
# Read the analytic data
data <- FinalAnalyticData

# Convert categorical variables to factors
data$chas_factor <- as.factor(data$chas_factor)
data$nox_cat <- as.factor(data$nox_cat)
data$dis_cat <- as.factor(data$dis_cat)
data$location <- as.factor(data$location)

# Smallest model: only location
model_small <- lm(medv ~ location, data = data)
print(round(summary(model_small)$coef, 3))
```
2. **full mode** contains all predictor variables

$$
\text{MEDV} = \beta_0 + \beta_1 \times \text{location} + \beta_2\times\text{chas_factor} + \beta_3\times \text{nox_cat}+\beta_3\times \text{dis_cat}
$$
```{r}
model_full <- lm(medv ~ chas_factor + nox_cat + dis_cat + location, data = data)
print(round(summary(model_full)$coef, 3))
```

3. **Auto Selected Model**

Several automatic model selection methods, including forward selection, backward elimination, stepwise selection, and best subset selection, can be used to search for a suitable candidate model. In this analysis, we illustrate the forward selection approach.

```{r}
# (3) Automatic variable selection algorithm

# Forward selection
model_forward <- step(lm(medv ~ 1, data = data),
                      scope = medv ~ chas_factor + nox_cat + dis_cat + location,
                      direction = "forward",
                      trace = FALSE)

print(round(summary(model_forward)$coef, 3))
```

## Residual Dianostic Analysis

The residual-based plots—namely, the residual plot, Q-Q plot, scale-location plot, and leverage plot—are summarized as follows.

```{r}

library(ggplot2)
library(dplyr)

# Diagnostic plots for the model_forward model

# Create diagnostic data set
residual_data <- data.frame(
  Fitted = fitted(model_forward),
  Residuals = residuals(model_forward),
  Standardized = rstandard(model_forward),
  Studentized = rstudent(model_forward)
)
my.theme <- function(){
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      axis.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
}
# Residuals vs Fitted plot
p1 <- ggplot(residual_data, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, method = "loess", color = "blue") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals") +
  my.theme()

# Q-Q plot
p2 <- ggplot(residual_data, aes(sample = Residuals)) +
  stat_qq(alpha = 0.6, color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  my.theme()

# Scale-Location plot
p3 <- ggplot(residual_data, aes(x = Fitted, y = sqrt(abs(Standardized)))) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(se = FALSE, method = "loess", color = "blue") +
  labs(title = "Scale-Location Plot",
       x = "Fitted Values",
       y = expression(sqrt("|Standardized Residuals|"))) +
  my.theme()

# Residuals vs Leverage plot
leverage <- hatvalues(model_forward)
cooksd <- cooks.distance(model_forward)

residual_data$Leverage <- leverage
residual_data$CookSD <- cooksd

p4 <- ggplot(residual_data, aes(x = Leverage, y = Studentized)) +
  geom_point(aes(size = CookSD), alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Leverage",
       x = "Leverage",
       y = "Studentized Residuals") +
  scale_size_continuous("Cook's Distance") +
  my.theme()

# Arrange plots
library(gridExtra)
grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```
The diagnostic plots suggest that the regression assumptions should be examined carefully. In particular, the Q-Q plot may indicate some departure from normality, while the residual and scale-location plots can be used to assess possible nonlinearity or heteroscedasticity. If strong skewness or unequal variance is present, a transformation of the response variable `medv` may be considered. Therefore, it is reasonable to further investigate whether a Box-Cox transformation can improve model fit.

## Box-Cox Transformation

We perform a Box-Cox transformation based on the candidate model `model_forward`.


```{r}
# (5) Box-Cox transformation

# Load MASS package
library(MASS)

# Perform Box-Cox transformation
boxcox_result <- boxcox(model_forward, lambda = seq(-2, 2, length = 100))

# Find optimal lambda
lambda_opt <- boxcox_result$x[which.max(boxcox_result$y)]
cat("Optimal lambda for Box-Cox transformation:", lambda_opt, "\n")
```
The Box-Cox result will be used to assess whether transforming medv is necessary. If the suggested $\lambda$ differs substantially from 1, we will consider an appropriate transformation of the response variable in the subsequent modeling step.

The Box-Cox curve indicates that $\lambda =0.5$ s a reasonable choice. This suggests using the square root of medv as the transformed response variable. The updated model is

$$
\sqrt{\text{MEDV}} = \beta_0 + \beta_1 \times \text{location} + \beta_2\times\text{chas_factor} + \beta_3\times \text{nox_cat}+\beta_3\times \text{dis_cat}
$$

Some of the key goodness-of-fit metrics for the transformed model are summarized as follows.

```{r}
# Create transformed response variable
if(lambda_opt != 0) {
  data$medv_trans <- (data$medv^lambda_opt - 1) / lambda_opt
} else {
  data$medv_trans <- log(data$medv)
}

# Fit model with transformed response
model_transformed <- lm(medv_trans ~ chas_factor + nox_cat + dis_cat + location, data = data)

# Compare original and transformed models
cat("\n=== Model Comparison ===\n\n",
    "Original model R-squared:", summary(model_forward)$r.squared, "\n\n",
    "Transformed model R-squared:", summary(model_transformed)$r.squared, "\n\n",
    "Original model AIC:", AIC(model_forward), "\n\n",
    "Transformed model AIC:", AIC(model_transformed), "\n")
```

The above $R^2$indicates only a limited improvement.

**Caution:** Because AIC is calculated from the log-likelihood, and the log-likelihood depends on the scale of the response variable, the AIC values from the original model and the transformed model are **not directly comparable**.

```{r}
# Create transformed response variable
data$medv_trans <- sqrt(data$medv)

# Fit transformed model
model_transformed <- lm(medv_trans ~ chas_factor + nox_cat + dis_cat + location, data = data)

# Diagnostic plots for transformed model
residual_data_trans <- data.frame(
  Fitted = fitted(model_transformed),
  Residuals = residuals(model_transformed),
  Standardized = rstandard(model_transformed)
)

# Diagnostic plots for transformed model
residual_data_trans <- data.frame(
  Fitted = fitted(model_transformed),
  Residuals = residuals(model_transformed),
  Standardized = rstandard(model_transformed)
)

p1_trans <- ggplot(residual_data_trans, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(se = FALSE, method = "loess", color = "blue") +
  labs(title = "Transformed Model: Residuals vs Fitted", 
       x = "Fitted Values", 
       y = "Residuals") +
  theme_minimal()

p2_trans <- ggplot(residual_data_trans, aes(sample = Residuals)) +
  stat_qq(alpha = 0.6, color = "steelblue", size = 3) +
  stat_qq_line(color = "red") +
  labs(title = "Transformed Model: Normal Q-Q Plot", 
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles") +
  theme_minimal()

grid.arrange(p1_trans, p2_trans, ncol = 2)
```

The diagnostic plots for the transformed model still suggest some departure from the normality assumption. Since the transformed model also reduces the direct interpretability of the response variable, we retain the original candidate model for reporting and interpretation. Although the residuals are not perfectly normal, the Boston housing dataset has a reasonably large sample size, so the reported coefficient tests are still expected to be approximately valid under large-sample theory.

# Summary of Optimal Model

A multiple linear regression model was estimated to explain median house value (`medv`) using selected neighborhood and location-related predictors. The final model includes `location`, `chas_factor`, `nox_cat`, and `dis_cat`. Overall, the model indicates that these engineered predictors are useful for explaining variation in Boston housing values.

**Key Predictors**:

* **Distance to Employment Centers (`dis_cat`)**: House values differ across distance categories, suggesting that neighborhood accessibility and proximity to employment centers are important determinants of housing value.

* **Nitric Oxides Concentration (`nox_cat`)**: Housing values vary across pollution categories, indicating that environmental quality is associated with property values. In general, areas with more favorable air quality tend to have higher median house values.

* **Charles River Indicator (`chas_factor`)**: Tracts bordering the Charles River may differ in median house value compared with those that do not, suggesting that this location-related factor captures an important neighborhood characteristic.

* **Location Type (`location`)**: The location grouping based on highway accessibility contributes to explaining differences in housing values, although its effect should be interpreted jointly with the other predictors in the model.

Overall, the model suggests that environmental conditions, accessibility, and neighborhood structure are all associated with median housing values in the Boston area. While the model does not explain all of the variability in `medv`, it provides a meaningful summary of how these selected factors relate to housing prices.









